# Regression

See https://leanpub.com/regmods

And https://github.com/bcaffo/courses/blob/master/07_RegressionModels/

## Sample Data

```{r message=FALSE, warning=FALSE}
library(UsingR)
library(dplyr)
library(ggplot2)

data(galton)
head(galton)
```

## Basic Idea

Assume we have a data set with two vectors X and Y, and we would like to find a model that uses X to predict Y.  In other words, we have Y = f(X).

To do basic linear regression, we are looking for a line of best fit (i.e. y = a + bx).  The two coefficients  that we are looking for (a,b) are the Y intercept and the slope.  To calculate these values, here's a few quick statistical formulas to refresh yourself

## The Emprical Variance (Stand Dev is its square root)
$$
S^2 = \frac{1}{n-1} \sum_{i=1}^n (X_i - \bar X)^2 
= \frac{1}{n-1} \left( \sum_{i=1}^n X_i^2 - n \bar X ^ 2 \right)
$$

## The Empirical Covariance and Correlation
Consider now when we have pairs of data, $(X_i, Y_i)$.  Their empirical covariance is 
$$
Cov(X, Y) = 
\frac{1}{n-1}\sum_{i=1}^n (X_i - \bar X) (Y_i - \bar Y)
= \frac{1}{n-1}\left( \sum_{i=1}^n X_i Y_i - n \bar X \bar Y\right)
$$
The correlation is defined is
$$
Cor(X, Y) = \frac{Cov(X, Y)}{S_x S_y}
$$

## Finding Intercept and Slope for Line of Best Fit
The line passes through the point $(\bar X, \bar Y$)

The slope of the regression line with $X$ as the outcome and $Y$ as the predictor is $Cor(Y, X) Sd(X)/ Sd(Y)$. 

Let's calculate it
```{r}
y <- galton$child
x <- galton$parent
beta1 <- cor(y, x) *  sd(y) / sd(x)
beta0 <- mean(y) - beta1 * mean(x)
c(beta0, beta1)
```

R has a function called lm() to do this for use (i.e. Linear Model)


```{r}
lm(y ~ x)
```

## Plotting with ggplot (Use geom_smooth with method=lm)

```{r}
freqData <- as.data.frame(table(galton$child, galton$parent))
names(freqData) <- c("child", "parent", "freq")
freqData$child <- as.numeric(as.character(freqData$child))
freqData$parent <- as.numeric(as.character(freqData$parent))

g <- ggplot(filter(freqData, freq > 0), aes(x = parent, y = child))
g <- g  + scale_size(range = c(2, 20), guide = "none" )
g <- g + geom_point(colour="grey50", aes(size = freq+20, show_guide = FALSE))
g <- g + geom_point(aes(colour=freq, size = freq))
g <- g + scale_colour_gradient(low = "lightblue", high="white")  
g <- g + geom_smooth(method="lm", formula=y~x)
g

```


