---
title: "Data Structures"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Table of Contents

1. [Table of Contents](#table-of-contents)
2. [Basics](#some-terminology-basics)
3. [Data frame basics](#data-frame-basics)
4. [Filter and Project](#filter-and-project)
5. [Adding Rows and Columns](#adding-rows-and-columns)
6. [Joining and Aggregating](#joining-and-aggregating)


### Some terminology basics

1. Types of **atomic vectors** (see [more](http://arrgh.tim-smith.us/atomic.html) on **vectors** from Tim Smith's page)
    - logical (may contain TRUE, FALSE, NA)
    - integer
    - double
    - complex (as in complex numbers; write as 0+0i)
    - character
    - raw (for bitstreams; printed in hex by default. Logical operators magically operate bitwise on these; they operate elementwise on all other vector types.)
    
2. **Arrays** include
    - **vectors** (i.e. 1 dimensional **arrays**)
    - **matrices** (i.e. 2 dimensional **arrays**)
    
4. Type of dimension labels:
    - `dimnames()` for **arrays**
    - `rownames()`, `colnames()` for **matrices** and **data frames**
    - `names()` for **atomic vectors** and **lists**
	
5. Coercing a **data frame**
    - From data frame to **array**: gets treated like 2d **matrices**
    - From data frame to **list** : gets treated like **list of vectors**

[top](#table-of-contents)

### Data Frame Basics

```{r data-frame-example}
df = data.frame( 
  col.1 = as.factor(c('a','b','a','b')), 
  col.2 = rnorm(1:4), 
  col.3 = 5:8 
)
print(df)

# Isolate a column i.e. a dimension, a sample variable
print(df$col.2)

# Isolate a row i.e. a record, a sample observation
print(df[1,])

# An abbreviated view of the data frame
str(df)
```
[top](#table-of-contents)

### Filter and Project

```{r filter-and-protect}

# Create copy of data frame after filtering out rows i.e. records, sample observations
subset(df, df$col.1=='a')

# You may filter and project a data frame with [,]
# From an SQL perspective, visualize it as dataframe[ WHERE, SELECT ]
#
df[ df$col.3 > 6, ] # Note 2nd param in [,] is blank to indicate all variables
```
[top](#table-of-contents)

### Adding Rows and Columns
```{r add-rows-and-cols}
# Adding a column (i.e. dimension, variable) is easy
df$col.4 = NA
print(df)

# Adding rows is easy too. First, we will make some new rows
df.more = data.frame( 
  col.1 = as.factor(c('a','b')), 
  col.2 = rnorm(1:2), 
  col.3 = 9:10,
  col.4 = rep("new",2)
)

# Now combine the rows, as in an SQL UNION
rbind(df, df.more)
```
[top](#table-of-contents)

### Joining and Aggregating

```{r pply}

# Create a reference table.  We will use this to do a key lookup based on col.1
df.ref = data.frame( 
  col.1 = as.factor(c('a','b')), 
  col.ref =c('AA','BB')
)
print(df.ref)

# Using the plyr library we can JOIN the reference data to the original data
library(plyr)
join(df, df.ref, by="col.1")

# We can also use the plyr library for GROUP BY operations on data projections
library(plyr)
ddply( df, .(col.1), summarize, sum.col.2 = sum(col.2), min.col.3 = min(col.3) )
```
[top](#table-of-contents)

