---
title: "Data Structures"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Table of Contents

1. [Table of Contents](#table-of-contents)
2. [Basics](#some-basics)
3. [Data Frames](#data-frame-basics)
    - [Get Rows and Columns](#get-rows-and-columns)
    - [Add Rows and Columns](#add-rows-and-columns)
    - [Change Columns](#change-columns)
    - [Filter and Project](#filter-and-project)
    - [Order By](#order-by)
    - [Warning with Brackets](#warning-with-brackets)
    - [Join and Aggregate](#join-and-aggregate)
4. [Lists](#list-basics)


### Some Basics

1. Types of **atomic vectors** (see [more](http://arrgh.tim-smith.us/atomic.html) on **vectors** from Tim Smith's page)
    - logical (may contain TRUE, FALSE, NA)
    - integer
    - double
    - complex (as in complex numbers; write as 0+0i)
    - character
    - raw (for bitstreams; printed in hex by default. Logical operators magically operate bitwise on these; they operate elementwise on all other vector types.)
    
2. **Arrays** include
    - **vectors** (i.e. 1 dimensional **arrays**)
    - **matrices** (i.e. 2 dimensional **arrays**)
    
4. Type of dimension labels:
    - `dimnames()` for **arrays**
    - `rownames()`, `colnames()` for **matrices** and **data frames**
    - `names()` for **atomic vectors** and **lists**
	
5. Coercing a **data frame**
    - From data frame to **array**: gets treated like 2d **matrices**
    - From data frame to **list** : gets treated like a **list of vectors**

[top](#table-of-contents)

### Data Frame Basics
```{r data-frame-example}
set.seed(1234) # Use same random data each time

# Here is a data frame
df = data.frame( 
  col.1 = as.factor(c('a','b','a','b')), 
  col.2 = rnorm(1:4), 
  col.3 = 5:8 
)
print(df)
# An abbreviated view of the data frame.  Good for investigation of newly read/created data
str(df)
```
[top](#table-of-contents)

#### Get Rows and Columns
```{r get-rows-and-cols}
# You may get/set the column names 
#   i.e. names of dimensions 
#   i.e. names of sample variables
colnames(df)
# Isolate a column, NOTE that this yields a vector (array of homogenous type)
print( df$col.2 )
# You may get/set the row names
#   i.e. names of records 
#   i.e. names of tuples 
#   i.e. names of sample observations
rownames(df)
# Isolate a row , NOTE that this yields a data.frame (with only one row)
print( df[1,] )
```
[top](#table-of-contents)

#### Add Rows and Columns
```{r add-rows-and-cols}
# Adding a column is easy
df$col.4 = NA
print(df)

# Adding rows is easy too. First, we will make some new rows
df.more = data.frame( 
  col.1 = as.factor(c('a','b')), 
  col.2 = rnorm(1:2), 
  col.3 = 9:10,
  col.4 = rep("new",2)
)

# Now combine the rows, as in an SQL UNION
rbind(df, df.more)
```
[top](#table-of-contents)

#### Change Columns
```{r change-columns}
# Currently df$col.1 is a factor vector. Lets change it to a character vector
df$col.1 = as.character(df$col.1)
str(df)
# Now change df$col.1 back to a vector using a different syntax
df = transform(df, col.1 = as.factor(col.1))
str(df)
```
[top](#table-of-contents)

#### Filter and Project
```{r filter-and-protect}

# Create filtered subset of rows
subset(df, df$col.1 == 'a')

# You may filter and project a data frame with [,]
# From an SQL perspective, visualize it as dataframe[ WHERE, SELECT ]
#
df[ df$col.1 == 'a', ] # Note 2nd param in [,] is blank to indicate all variables
# Here it is again but with specific columns being projected
df[ df$col.1 == 'a', c("col.1", "col.3") ]
```
[top](#table-of-contents)

#### Order By
```{r order-by}
# Recall you may access a row by its name, which by default is an integer sequence: 1, 2, ..., nrow
print( df[1,] )
# You may also access rows in order of your choice
df[ nrow(df):1, ]
# The order() function in base R takes column names (use '-' for descending)
df[ order(df$col.1, -df$col.2), ]
```
[top](#table-of-contents)

#### Warning with Brackets
```{r brackets}
# Be careful when using [] with dataframe instead of [,]
#   this specifies what column(s) to project

# From an SQL perspective, visualize it as dataframe[ SELECT ] 
df[1]
# May also use a column name as a key
df["col.1"]
```
[top](#table-of-contents)

#### Join and Aggregate

```{r pply}

# Create a reference table.  We will use this to do a key lookup based on col.1
df.ref = data.frame( 
  col.1 = as.factor(c('a','b')), 
  col.ref1 = c('AA','BB'),
  col.ref2 = c('aaa','bbb')
)
print(df.ref)

# Using the plyr library we can JOIN the reference data to the original data
library(plyr)
join(df, df.ref, by="col.1")

# We can also use the plyr library for GROUP BY operations on data projections
library(plyr)
ddply( df, .(col.1), summarize, sum.col.2 = sum(col.2), min.col.3 = min(col.3) )
```
[top](#table-of-contents)

### List Basics

```{r list-basics}
# Create a list.  The '.' function from the plyr library is handy for this.
library(plyr)
x = .(1, 2, "three", as.factor(4))
str(x)
# Subset the list
x[1:3]
# Subset the list to only include a specific element
x[1]
# Obtain element from list
x[[1]] # equivalent to unlist(x[1]), see below
# Convert from list subset to vector of contained elements (assuming consistent element types)
unlist( x[1:2] )
# If a subset query matches no elements, then a list with NULL as its only element is returned
x[5000]
# However trying to access an element that does not exist is an error state
tryCatch({
    x[[5000]]
  }
  ,error=function(cond) {
    print(cond)
  }
)
```
[top](#table-of-contents)





